---
title: "Stat 5201 - Assignment 2"
author: "Bailey Perry"
date: "October 2, 2017"
output: pdf_document
---
# Textbook problems:
General information -- Simple Probability Samples Exercises.

## 1. Chapter 2 - Problem 1:
Let N=6 and n=3, and assume all population values are known. The book has all of the tables laid out for the two sampling plans.

###a) What is the value of ybar(U)?
```{r}
ybaru <- (98+102+154+133+190+175)/6
ybaru
```
ybar(U) = sum(yi)/N = 142

###b) Let ybar be the mean of the sample plans. Find E[ybar], Var(ybar), Bias(ybar), and MSE(ybar).

E[ybar] for each sampling plan:
```{r}
plan1mu <- (mean(c(98,154,190))*1/8)+(mean(c(98,154,175))*1/8)+(mean(c(98,133,190))*1/8)+(mean(c(98,133,175))*1/8)+(mean(c(102,154,190))*1/8)+(mean(c(102,154,175))*1/8)+(mean(c(102,133,190))*1/8)+(mean(c(102,133,175))*1/8)
plan1mu
```

```{r}
plan2mu <- (mean(c(98,133,175))*1/4)+(mean(c(102,154,175))*1/2)+(mean(c(98,154,190))*1/4)
plan2mu
```
Var(ybar) for each sampling plan:
This is calculated by taking the product of (sample means-ybar(U))^2 and multiplying that by their associated probabilities, and then summing all of those values. This is seen below.
```{r}
plan1var <- (((147.33-plan1mu)^2)*1/8)+(((142.33-plan1mu)^2)*1/8)+(((140.33-plan1mu)^2)*1/8)+(((135.33-plan1mu)^2)*1/8)+(((148.67-plan1mu)^2)*1/8)+(((143.67-plan1mu)^2)*1/8)+(((141.67-plan1mu)^2)*1/8)+(((136.67-plan1mu)^2)*1/8)
plan1var
```
```{r}
plan2var <- (((135.33-plan2mu)^2)*1/4)+(((143.67-plan2mu)^2)*1/2)+(((147.33-plan2mu)^2)*1/4)
plan2var
```
Bias(ybar) for each sampling plan:
Bias(ybar)=E[ybar]-ybar(U), all values we have from above.
```{r}
p1bias <- plan1mu-ybaru
p1bias
```
```{r}
p2bias <- plan2mu-ybaru
p2bias
```
MSE(ybar) for each sampling plan:
MSE(ybar)=Var(ybar)+(Bias(ybar)^2)
```{r}
p1mse <- plan1var+(p1bias^2)
p1mse
```
```{r}
p2mse <- plan2var+(p2bias^2)
p2mse
```
###c) Which sampling plan is better? Why?
From all of the output above, we can clearly see that Plan 1 is better in this case. Not only is it closer to the true population mean, but it also has a smaller variance. More importantly, the mean squared error is smaller for plan 1, as is the bias value for plan 1. All of these point to plan 1 doing a better job of generalizing to the population, and thus it is the better sampling choice.

## 2. Chapter 2 - Problem 2:

###a) Find the probability of selection pi(i) for each unit i.
1 is found in sample one and three, thus pi(1)=(1/8)+(1/4)= 3/8

2 is found in sample two and four, thus pi(2)=(1/4)+(3/8)=5/8

3 is found in sample one and two, thus pi(3)=(1/8)+(1/4)=3/8

4 is found in sample three, four, and five, thus pi(4)=(1/8)+(1/8)+(3/8)=5/8

5 is found in sample one and five, thus pi(5)=(1/8)+(1/8)=2/8

6 is found in sample one, three, and four, thus pi(6)=(1/8)+(1/8)+(3/8)=5/8

7 is found in sample two and five, thus pi(7)=(1/4)+(1/8)=3/8

8 is found in sample two, three, four, and five, thus pi(8)=1-(1/8)=7/8

###b) What is the sampling distribution of t(est)=8ybar?
```{r}
#First find the means for each of the 5 samples
s1ybar <- (1+4+7+7)/4
s2ybar <- (2+4+7+8)/4
s3ybar <- (1+4+7+8)/4
s4ybar <- (2+4+7+8)/4
s5ybar <- (4+7+7+8)/4

#Next, find their value for t=8(ybar)
t1 <- s1ybar*8
t1
t2 <- s2ybar*8
t2
t3 <- s3ybar*8
t3
t4 <- s4ybar*8
t4
t5 <- s5ybar*8
t5
```
From the above we notice that t2 and t4 both are equal to 42, thus we need to add their respective probabilities to find the overall distribution.

t(i), P(t(i))

38, P=1/8

40, P=1/8

42, P=5/8

52, P=1/8

## 3. Chapter 2 - Problem 6:
University has 807 faculty members. Each one has a certain number of refereed publications. This problem is based on a SRS of 50 faculty members.

###a) Plot the data using a histogram. Describe the shape of the data.
```{r}
x <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
y <- c(28, 4, 3, 4, 4, 2, 1, 0, 2, 1, 1)
z <- rep.int(x,times=y)
hist(z, nclass=11, xlim=c(0,12), ylim=c(0,30), right=FALSE, main="Refereed Publications by Faculty Members", xlab="Papers")
```
The shape of the data is right-skewed. Most of the professors have few to no refereed papers, and thus there is a high concentration in these low values. The value for zero papers is also much higher than any of the others.

###b) Estimate the mean number of publications per faculty member, and give the SE for your estimate.
```{r}
estybar <- ((28/50)*0)+((4/50)*1)+((3/50)*2)+((4/50)*3)+((4/50)*4)+((2/50)*5)+((1/50)*6)+((0/50)*7)+((2/50)*8)+((1/50)*9)+((1/50)*10)
estybar
```
From this, we see that the estimated mean number of publications per faculty member is 1.78 papers.

Next we calculate the estimated variance, to then obtain standard error (the square root of the estimate). 
```{r}
estvar <- var(z)
estvar

#Thus we know that standard error=sqrt((estvar/n)*(1-n/N))
paperse <- sqrt((estvar/50)*(1-(50/807)))
paperse
```
###c) Do you think that ybar from (b) will be normally distributed? Why or why not?
I do not believe that the mean estimate from (b) will be normally distributed. This is because the data is extremely skewed and has the largest proportion at 0 papers, and it does not seem to have any indications of normality. Although, it is important to consider that a larger sample size may allow the use of the central limit theorem and thus introduce normality to those plots.

###d) Estimate the proportion of faculty members with no publications and give a 95% CI.
```{r}
#Sample proportion
prop <- 28/50
prop

#Standard error of the proportion - use formula below
#SEp = sqrt[p*( 1 - p )/ n-1] * sqrt[(1-(n/N))]
sep <- sqrt((prop*(1-prop)/49))*sqrt(1-(50/807))
sep

#Now we need to find the critical value for the formula
t <- qt(0.975, df=49)
t
```
```{r}
upper1 <- prop + sep*t
lower1 <- prop - sep*t
lower1; upper1
```
Thus, a 95% CI is the sample proportion +/- the critical value times the standard error of the sample proportion. In numbers that is: [0.422, 0.698]. Therefore the estimated proportion of faculty with no publications is 56%, and it falls within the range 95% of the time. 

## 4. Chapter 2 - Problem 7:

###a) Find a 95% CI for the percentage of entries that come from the South.
```{r}
#Sample proportion
prop1 <-175/1000 
prop1

#Standard error of the proportion - use formula below
#SEp = sqrt[p*( 1 - p )/ n]
# Since the sample is sufficiently large, we use the above
# adjusted formula
sep1 <- sqrt((prop1*(1-prop1)/1000))
sep1

#Now we need to find the critical value for the formula
t1 <- qt(0.975, df=999)
t1
```

```{r}
upper2 <- prop1 + sep1*t1
lower2 <- prop1 - sep1*t1
lower2; upper2
```
Thus, a 95% CI is the sample proportion +/- the critical value times the standard error of the sample proportion. In numbers that is: [0.1514, 0.1986]. Therefore the estimated percentage of entries from the South is 17.5%, and it falls within the given range 95% of the time.

###b) Is there evidence from your CI that the percentage of entries from the South differs from the percentage of persons living in the South?
Yes, there is evidence because the value 30.9% is not in the confidence interval. This suggests that there is a difference between the percentage of entries from the South, and the percentage of people living there.

## 5. Consider a population of N=580 patients of a clinic. A quantity of interest is the percentage of patients that overdue for a vaccination.

###(a) What sample size in a SRS(without replacement) would be necessary to estimate the proportion with 95% confidence and margin of error 0.10.
```{r}
#First find n0
#We use p=0.5 since that maximizes the value and p isn't given
n0 <- (((1.96^2)*0.5*(1-0.5))/0.10^2)
n0

#Then we can find n, set N=580, given above
n <- n0/(1+(n0/580))
n
```
The necessary sample size looks to be n=83, when rounded up to the nearest whole number. This value will meet the requirements.

###(b) Suppose a SRSWR(with replacement) of size 120, yielded 27 whom were not overdue for vaccination. Give a 95% CI for the proportion of children not overdue for vaccination.
```{r}
prop2 <- 27/120
prop2

#Standard error of the proportion - use formula below
#SEp = sqrt[p*( 1 - p )/ n-1] * sqrt[(1-(n/N))]
sep2 <- sqrt((prop2*(1-prop2)/119))*sqrt(1-(120/580))
sep2

#Now we need to find the critical value for the formula
t2 <- qt(0.975, df=119)
t2
```

```{r}
upper3 <- prop2 + sep2*t2
lower3 <- prop2 - sep2*t2
lower3; upper3
```
Thus, a 95% CI is the sample proportion +/- the critical value times the standard error of the sample proportion. In numbers that is: [0.1575, 0.2925]. Therefore the estimated proportion of children not overdue for vaccination was 22.5%, and it falls within the above range 95% of the time.

## 6. In R or your favorite computing environment, write a program that can generate a simple random sample without replacement and calculate the usual 95% confidence interval for the population mean. 
```{r}
set.seed(123)
pop <- rnorm(500)
mu <- mean(pop)
pvar <- var(pop)
confup <- mu + sqrt(pvar)*1.96
conflow <- mu - sqrt(pvar)*1.96
conflow; confup
```
###Construct a population of size 500 for which the usual interval based on a sample of size 50 covers the true mean less than 50% of the time. Show this is true by considering the results from 500 random samples.
```{r}
samp <- sample(pop, 50)
sampmu <- mean(samp)
sampvar <- var(samp)
#Need new z critical value for 50% confidence
zstat <- qnorm(0.75) 
sconfup <- sampmu + sqrt(sampvar)*zstat
sconflow <- sampmu - sqrt(sampvar)*zstat
sconflow; sconfup
```
```{r}
set.seed(123)
pop <- rnorm(500)
meanest1<-function(y,n,R){
  N<-length(y)
  popmn<-mean(y)
  ans<-rep(0,4)

    for(i in 1:R){
      smp<-sample(1:N,n)
      ysmp<-y[smp]
      estmn<-mean(ysmp)
      abserr<-abs(estmn - popmn)
      estvr<-(1 - n/N)*var(ysmp)/n
      lwbd<-estmn - qnorm(0.75)*sqrt(estvr)
      upbd<-estmn + qnorm(0.75)*sqrt(estvr)
      if(lwbd <= popmn & popmn <= upbd) {cov<-1}
      else {cov<-0}
      ans<-ans + c(estmn,abserr,2*qnorm(0.75)*sqrt(estvr),cov)
    }
    ans<-round(ans/R,digits=3)
    cat("  est ", "abserr  ", "length ", "freq of coverage","\n")
    print(ans)
    return(date())
  }
y<-pop
meanest1(y,50,500)
```

###For extra credit, construct a population and give a mathematical argument that shows the ususal interval will cover only 10% of the time.
```{r}
set.seed(123)
pop1 <- rnorm(500)
```

```{r}
meanest1<-function(y,n,R){
  N<-length(y)
  popmn<-mean(y)
  ans<-rep(0,4)

    for(i in 1:R){
      smp<-sample(1:N,n)
      ysmp<-y[smp]
      estmn<-mean(ysmp)
      abserr<-abs(estmn - popmn)
      estvr<-(1 - n/N)*var(ysmp)/n
      lwbd<-estmn - qnorm(0.55)*sqrt(estvr)
      upbd<-estmn + qnorm(0.55)*sqrt(estvr)
      if(lwbd <= popmn & popmn <= upbd) {cov<-1}
      else {cov<-0}
      ans<-ans + c(estmn,abserr,2*qnorm(0.55)*sqrt(estvr),cov)
    }
    ans<-round(ans/R,digits=3)
    cat("  est ", "abserr  ", "length ", "freq of coverage","\n")
    print(ans)
    return(date())
  }
y<-pop1
meanest1(y,50,500)
```
